Total Team Spending and Profitability Results in a Higher League Ranking in the English Premier League

### 2024-04-24 paper/paper.qmd
Final commits.

Established in 1992, the English Premier League comprises twenty clubs that compete annually, embarking on a 38 week season each August, each striving not only for the championship but also to avoid relegation. The English Premier League is the wealthiest football league globally, producing a total revenue of €10.5 billion as of the end of the 2022/2023 season [@EPLHistory]. Since the league began in 1992, only seven teams have come out victorious, with three of the seven teams only winning once (Leicester City 2015/16, Blackburn Rovers 1994/95 and Liverpool 2019/20) [@PremierLeagueExplained].

@@

The analysis dataset used for this paper is comprised of five datasets that were compiled during the data cleaning process. The first dataset, @tbl-1, contains every teams average home matchday attendance for the 2023-2024 season, which was collected from @footballwebpages2023premierleague. The second dataset, @tbl-2, contains every teams current market value, which was collected from @transfermarkt2023premierleague. The third dataset, @tbl-3, contains the current total payroll for each team in the English Premier League, and was collected from @spotrac2023eplpayroll. The forth dataset, @tbl-4, contains the total amount spent in transfer fees prior to the start of the 2023-2024 season for each team in the English Premier League, and was collected from @spotrac2023epltransferfees. The final dataset, @tbl-5, contains the total points earned by each team in the English Premier League during the 2023-2024 as of Matchday 33, and was retrieved from @transfermarkt2023premierleaguematchday30.

@@

# Assigning new column names to the dataframe
colnames(total_payroll) <- c("Team", "Active Players", "Forwards", "Midfielders", "Defensmen", "Goalkeepers", "Total")

@@

Furthermore, to streamline the dataset and focus the analysis on key variables of interest, certain columns deemed extraneous to the core analytical objectives were removed. These included detailed payroll information per position (`forwards`, `midfielders`, `defensmen`, `goalkeepers`) and the column `active.players`. The removal of these columns served to reduce the dataset's complexity, facilitating a more focused and manageable analysis of the relationships between team performance metrics and financial expenditures.

@@

From here I began examining the relationships between English Premier League teams' league performance (expressed in points accumulated) and various financial indicators during the 2023-2024 season. The following four graphs (@fig-pts_vs_average_home_matchday_attendance, @fig-pts_vs_total_wage_bill, @fig-pts_vs_market_value, and @fig-pts_vs_transfer_fees) presents a scatter plot paired with a linear regression trend line to evaluate the correlation between league points and one of the four key economic metrics presented in @tbl-6.

@@

In @fig-pts_vs_average_home_matchday_attendance, @fig-pts_vs_total_wage_bill, @fig-pts_vs_market_value, and @fig-pts_vs_transfer_fees, each point point on the graph represents a team, plotted with their total points on the x-axis and their average home attendance on the y-axis. The trend line in @fig-pts_vs_average_home_matchday_attendance, added using linear regression analysis, suggests a correlation between team performance and matchday turnout. The correlation coefficient of 0.683116662462212 indicates a positive association with moderate correlation, where teams with higher league points tend to have greater average attendance. @fig-pts_vs_total_wage_bill linear regression trend line suggests a correlation between a team’s financial expenditure on payroll and their success in the league. A correlation coefficient of 0.788682340622421 reflects a positive relationship, implying that teams with higher payrolls may be more likely to accumulate greater league points. @fig-pts_vs_market_value demonstrates a pronounced upward trend represented by the linear regression line, with a correlation coefficient of 0.88165690596531, indicative of a strong positive relationship. This suggests that teams with higher market values tend to amass more league points, underlining a potential link between financial strength and on-field success. @fig-pts_vs_transfer_fees trend line indicates a moderate positive correlation, with a correlation coefficient of approximately 0.449012469460392. This suggests that higher league points tend to coincide with increased transfer spending, although the relationship is not strongly linear.

@@

The linear regression model given by Equation (2), derived from the foundational principles of linear regression outlined in Equation (1), provides a transparent mechanism to quantify the impact of various factors on the total points accumulated by EPL teams (a direct correlation to team rank in league standings). The model will aid in gaining insights into how attendance, total payroll, and market value could affect performance metrics. The model is operationalized using the `lm` function in `R`. A strong positive relationship is expected between the dependent variable (`Points`) and three independent variables (`Home Matchday Attendance`, `Total Payroll`, and `Market Value`). The higher the three independent variables are, the more points there should be accumulated by the team.

@@

The results from the linear regression model indicate a strong relationship between points earned by English Premier League teams and variables such as average home matchday attendance, total wage bill, and market value. The model explains a significant portion of the variability in points, with an $R^2$ value of 0.8057, suggesting that about 80.57% of the variation in team points can be accounted for by these three predictors. Market value is particularly impactful, as indicated by a statistically significant positive coefficient ($p = 0.00151$), suggesting that higher market values are associated with higher points. Conversely, the coefficient for the total wage bill is negative, though not statistically significant, hinting that higher wage bills might not necessarily translate into better performance. This is somewhat surprising, however, results could stem from market value being such a strong predictor that is overshadows the impactfulness of total wage bill. Attendance shows a positive but non-significant impact on points. The model's residuals indicate some deviation from perfect prediction, with values ranging from -10.81 to 15.96, suggesting variability that the model doesn’t capture. Overall, the significant $p$-value for the model (6.159e-06) confirms the effectiveness of these predictors in forecasting league points and therefore the confirmation to move forward in the analysis.

@@

@tbl-9 presents a comparative analysis of the rankings of English Premier League teams, juxtaposing their actual standings against their predicted standings. The prediction is derived from Equation (2) that incorporates variables such as home attendance, total payroll, and market value. Each row in the table features a rank number, followed by the actual team standing corresponding to that rank, and the predicted team standing that would occupy the same rank based on the model's calculations. This layout allows for a direct comparison of where teams are predicted to stand based on certain metrics versus where they actually stand in the league, highlighting the predictive model's accuracy and identifying discrepancies between expected and actual performances. 
```{r}
#| label: tbl-9
#| tbl-cap: Actual Rank vs Predicted Rank Based on Home Attendance, Total Payroll, and Market Value for English Premier League Teams
#| echo: false
#| warning: false

@@

The application of predictive analytics football, is an expanding field that can provide a competitive edge to clubs. This study aimed to correlate financial factors with team success highlighting the potential of data-driven decision-making in enhancing a team's competitive performance. The significant correlation between a team’s market value and league points serves as evidence of the tangible benefits that can be gained from predictive analytics

The data presented in @sec-results indicates a general correlation between the predicted points, which take into account factors such as home attendance, payroll, and market value, and the actual points achieved by the teams in the English Premier League. As of Matchday 33, Most teams appear to be performing within a close range of their predicted outcomes, suggesting that the model used for predictions is fairly accurate or that most teams are performing as expected given their financial inputs.

@@

This outlier status of Chelsea F.C. draws attention and could prompt stakeholders to conduct a deeper analysis into why the club is not meeting the expectations set by its financial prowess. It raises questions about the efficacy of the investments made by the club and whether financial muscle is translating into performance on the pitch as anticipated.

Chelsea F.C.'s lower-than-expected league points despite its significant financial outlay exemplifies the multifaceted nature of football performance. The discrepancy between predicted and actual points for Chelsea F.C. is one of the largest among all teams, prompting a closer look at factors beyond the economic predictors used in this study. It serves as a reminder that team cohesion, tactical decisions, and possibly non-quantifiable elements such as team spirit and injuries can heavily influence results. As such, Chelsea F.C.'s example could be a launching pad for exploring the non-economic dimensions that interplay with financial resources to shape the fortunes of a football club.

@@

## Weaknesses and next steps

@@

The sample size, constrained to the current teams in the Premier League for a single season, poses a further limitation. The relatively small dataset may not capture the full variability and could potentially overestimate or underestimate the effects of the predictors. Larger datasets that span multiple seasons and include teams from different tiers of English football might provide a more comprehensive view of the financial dynamics at play.

The results of the current analysis could lead to an intriguing dimension of financial investment in European football as the next steps: the potential long-term benefits of allocating financial resources to youth development programs. While this analysis primarily examines current and direct financial factors of team success, there is growing evidence to suggest that investments in youth academies and development programs can be equally pivotal for sustained success.

>>>>

Established in 1992, the English Premier League comprises twenty clubs that compete annually, embarking on a 38 week season, each striving not only for the championship but also to avoid relegation. The English Premier League is the wealthiest football league globally, producing a total revenue of €10.5 billion as of the end of the 2022/2023 season [@EPLHistory]. Since the league began in 1992, only seven teams have come out victorious, with three of the seven teams only winning once (Leicester City 2015/16, Blackburn Rovers 1994/95 and Liverpool 2019/20) [@PremierLeagueExplained].

@@

The analysis dataset used for this paper is comprised of five datasets that were compiled during the data cleaning process. The first dataset, @tbl-1, contains every teams average home matchday attendance for the 2023-2024 season, which was collected from @footballwebpages2023premierleague. The second dataset, @tbl-2, contains every teams current market value, which was collected from @transfermarkt2023premierleague. The third dataset, @tbl-3, contains the current total payroll for each team in the English Premier League, and was collected from @spotrac2023eplpayroll. The fourth dataset, @tbl-4, contains the total amount spent in transfer fees prior to the start of the 2023-2024 season for each team in the English Premier League, and was collected from @spotrac2023epltransferfees. The final dataset, @tbl-5, contains the total points earned by each team in the English Premier League during the 2023-2024 as of Matchday 33, and was retrieved from @transfermarkt2023premierleaguematchday30.

@@

# Assigning new column names to the dataframe
colnames(total_payroll) <- c("Team", "Active Players", "Forwards", "Midfielders", "Defensemen", "Goalkeepers", "Total")

@@

Furthermore, to streamline the dataset and focus the analysis on key variables of interest, certain columns deemed extraneous to the core analytical objectives were removed. These included detailed payroll information per position (`forwards`, `midfielders`, `defensemen`, `goalkeepers`) and the column `active.players`. The removal of these columns served to reduce the dataset's complexity, facilitating a more focused and manageable analysis of the relationships between team performance metrics and financial expenditures.

@@

From here I began examining the relationships between English Premier League teams' league performance (expressed in points accumulated) and various financial indicators during the 2023-2024 season. The following four graphs (@fig-pts_vs_average_home_matchday_attendance, @fig-pts_vs_total_wage_bill, @fig-pts_vs_market_value, and @fig-pts_vs_transfer_fees) present a scatter plot paired with a linear regression trend line to evaluate the correlation between league points and each key economic metric presented in @tbl-6.

@@

In @fig-pts_vs_average_home_matchday_attendance, @fig-pts_vs_total_wage_bill, @fig-pts_vs_market_value, and @fig-pts_vs_transfer_fees, each point point on the graph represents a team, plotted with their total points on the x-axis and their average home attendance on the y-axis. The trend line in @fig-pts_vs_average_home_matchday_attendance demonstrates a positive correlation between team performance and matchday turnout. The correlation coefficient of 0.683116662462212 indicates a positive association with moderate correlation, where teams with higher league points tend to have a higher home matchday attendance. @fig-pts_vs_total_wage_bill linear regression trend line suggests a correlation between a team’s financial expenditure on payroll and their success in the league. A correlation coefficient of 0.788682340622421 reflects a positive relationship, implying that teams with higher payrolls may be more likely to accumulate greater league points. @fig-pts_vs_market_value demonstrates a pronounced upward trend represented by the linear regression line, with a correlation coefficient of 0.88165690596531, indicative of a strong positive relationship. This suggests that teams with higher market values tend to amass more league points, underlining a potential link between financial strength and on-field success. @fig-pts_vs_transfer_fees trend line indicates a moderate positive correlation, with a correlation coefficient of approximately 0.449012469460392. This suggests that higher league points tend to coincide with increased transfer spending, although the relationship is not strongly linear.

@@

The linear regression model given by Equation (2), derived from the foundational principles of linear regression outlined in Equation (1), provides a transparent mechanism to quantify the impact of various economic factors on the total points accumulated by EPL teams (a direct correlation to team rank in league standings). The model will aid in gaining insights into how attendance, total payroll, and market value could affect performance metrics. The model is operationalized using the `lm` function in `R`. A strong positive relationship is expected between the dependent variable (`Points`) and three independent variables (`Home Matchday Attendance`, `Total Payroll`, and `Market Value`). The higher the three independent variables are, the more points that should be accumulated by a team.

@@

The results from the linear regression model indicate a strong relationship between points earned by English Premier League teams and variables such as average home matchday attendance, total wage bill, and market value. The model explains a significant portion of the variability in points, with an $R^2$ value of 0.8057, suggesting that about 80.57% of the variation in team points can be accounted for by these three predictors. Market value is particularly impactful, as indicated by a statistically significant positive coefficient ($p = 0.00151$), suggesting that higher market values are associated with higher points. Conversely, the coefficient for the total wage bill is negative, though not statistically significant, hinting that higher wage bills might not necessarily translate into better performance. This is somewhat surprising, however, results could stem from market value being such a strong predictor that it is overshadowing the impactfulness of total wage bill. Attendance shows a positive but non-significant impact on points. The model's residuals indicate some deviation from perfect prediction, with values ranging from -10.81 to 15.96, suggesting variability that the model doesn’t capture. Overall, the significant $p$-value for the model (6.159e-06) confirms the effectiveness of these predictors in forecasting league points and therefore the confirmation to move forward in the analysis.

@@

@tbl-9 presents a comparative analysis of the rankings of English Premier League teams, juxtaposing their actual standings against their predicted standings. The prediction is derived from Equation (2) that incorporates variables such as home attendance, total payroll, and market value. Each row in the table features a rank number, followed by the actual team standing corresponding to that rank, and the predicted team standing that would occupy the same rank based on the model's calculations. This layout allows for a direct comparison of where teams are predicted to stand based on certain metrics versus where they actually stand in the league, highlighting the model's accuracy and identifying discrepancies between expected and actual performances. 
```{r}
#| label: tbl-9
#| tbl-cap: Actual Rank vs Predicted Rank Based on Home Attendance, Total Payroll, and Market Value for English Premier League Teams
#| echo: false
#| warning: false

@@

The application of predictive analytics in football is an expanding field that can provide a competitive edge to clubs. This study aimed to correlate financial factors with team success highlighting the potential of data-driven decision-making in enhancing a team's competitive performance. The significant correlation between a team’s market value and league points serves as evidence of the tangible benefits that can be gained from predictive analytics and understanding the significance of team wealth in professional sports.

The data presented in @sec-results indicates a strong positive correlation between the predicted points, which take into account factors such as home attendance, payroll, and market value, and the actual points achieved by the teams in the English Premier League. As of Matchday 33, Most teams appear to be performing within a close range of their predicted outcomes, suggesting that the model used for predictions is fairly accurate or that most teams are performing as expected given their financial inputs.

@@

This outlier status of Chelsea F.C. draws attention and could prompt stakeholders to conduct a deeper analysis into why the club is not meeting the expectations set by its financial prowess. It raises questions about the efficacy of the investments made by the club and whether financial power is translating into performance on the pitch as anticipated.

Chelsea F.C.'s lower-than-expected league points despite its significant financial outlay exemplifies the multifaceted nature of football performance. The discrepancy between predicted and actual points for Chelsea F.C. is one of the largest among all teams, prompting a closer look at factors beyond the economic predictors used in this study. It serves as a reminder that team cohesion, tactical decisions, and possibly non-quantifiable elements such as team spirit and injuries can heavily influence results. As such, Chelsea F.C.'s example could be a launching pad for exploring the non-economic dimensions that interplay with financial resources to shape the success of a football club.

@@

## Weaknesses and Next Steps

@@

The sample size, constrained to the current teams in the Premier League for a single season, poses a further limitation. The relatively small dataset may not capture the full variability and could potentially overestimate or underestimate the effects of the predictors. Larger datasets that span perhaps all four division/tiers of English football might provide a more comprehensive view of the financial dynamics at play.

This report primarily focuses on the immediate financial aspects of team success. However, there is increasing evidence that investments in English Premier League teams youth academies and development programs are crucial for long-term success. These potential benefits make a strong case for channeling financial resources into nurturing young talent as a strategic move for long term success within a club. Next steps could include analyzing the investment strategies and financial power present in youth development programs and could add a new dimension to understanding English Premier League teams success.
